---
output: html_document
---

## Computational environment in R

We set up a computational environment for spatial analysis in R using the `groundhog` package.
This package allows you to specify a date at which the analysis was developed and a list of packages required.
From there, groundhog checks for a version of R appropriate for that date and installs the latest versions of the required packages as of that date.

This is particularly important in the context of spatial data science because the packages we use have a history of major version upgrades without backward capability.
For example, version 4.0 of the `tmap` package for thematic mapping cannot be used to run all code written for version 3: most of the code will work with warnings, but not all.

In this book, we use:

- `here` to manage reproducible file paths
- `tidyverse` for data wrangling and graphing
- `sf` for vector data and algorithms aligned with the *simple features* standards
- `tmap` for cartography
- `lwgeom` for minimum bounding circles and geodesic calculations

The code block below may need to be run multiple times to get the full environment working.
You may also need to respond to prompts and warnings output to the console, including:

- `ok` for permission to write groundhog files to the groundhog folder
- install the correct version of R for the time given and start an R session with that version
- restart the R session and re-run the code block

```{r message=FALSE, warning=FALSE}
packages <- c("here", "tidyverse", "sf", "tmap", "lwgeom", "kableExtra")

# force all conflicts to become errors
# if you load dplyr and use filter(), R has to guess whether you mean dplyr::filter() or stats::filter()
# the conflicted package forces you to be explicit about this
# disable at your own peril
# https://conflicted.r-lib.org/
require(conflicted)

# load and install required packages
# https://groundhogr.com/
if (!require(groundhog)) {
  install.packages("groundhog")
  require(groundhog)
}

# this date will be used to determine the versions of R and your packages
# it is best practice to keep R and its packages up to date
groundhog.day <- "2025-02-01"

# you may want to use set.groundhog.folder() to choose where to install packages under groundhog
# the folder should be on the same drive as your R installation, and should not be synced to the cloud

# this replaces any library() or require() calls
groundhog.library(packages, groundhog.day)

# set up default knitr parameters: https://yihui.org/knitr/options/
knitr::opts_chunk$set(
  echo = TRUE,
  fig.retina = 4,
  fig.width = 8,
  fig.path = paste0(here("results", "figures"), "/")
)
```

Record the established computational environment for R.

```{r eval=FALSE}
sessionInfo()

# alternatively, use devtools::session_info() for better results
writeLines(
  capture.output(sessionInfo()),
  here("procedure", "environment", paste0("r-environment-", Sys.Date(), ".txt"))
)

knitr::write_bib(c(packages, c("knitr", "bookdown", "rmarkdown")), 
                 here('packages.bib'))
```

## Hello, Central Falls world

Let's read a reference point grid from the geopackage `cf_given` in the layer `pointgrid`.

As we read the data in, we'll recalculate x- and y- offsets for label placement in `tmap`.  
Those offsets worked well to place the grid reference letters and numbers just outside the grid in QGIS layouts, but they need to be adjusted to achieve the same placements in `tmap`.

```{r}
pointgrid <- st_read(here("data", "raw", "public", "cf_given.gpkg"), 
                     layer="pointgrid") |> 
  mutate(xOff = xOff / 4,
         yOff = yOff / -4,
         order = ifelse(order == 0, 0.5, order),
         label = ifelse(is.na(label), "", label))
```
As we load the data, *sf* reports some basic geographic metadata about the layer, including the geometry type, bounding box (geographic extent), and coordinate reference system.

Let's display the grid.

We make the `bbox` equal to a buffered `pointgrid` to add some margin space to the map figure.

```{r}
pointgrid_map <- pointgrid |> 
  tm_shape(bbox = st_buffer(pointgrid, 250)) +
  tm_symbols(col_alpha = 0.5,
             fill_alpha = 0,
             size = "order",
             size.legend = tm_legend_hide()) +
  tm_text("label",
          ymod="yOff",
          xmod="xOff",
          col="grey50",
          fontfamily = "mono")

pointgrid_map
```

Let's read a layer of Central Falls `zones`, the local geographic unit of enumeration and neighborhood governance.

```{r}
zones <- st_read(here("data", "raw", "public", "cf_digitized.gpkg"), 
                     layer="zones")
```

Let's inspect the `zones` attribute table.

```{r}
zones |> st_drop_geometry() |> knitr::kable()
```

Notice how the `id` column has been converted into row labels.
There are four other columns:

- `zone` is character string with the zone name based on compass direction.
- `pop` is an integer count of the total population
- `minority` is an integer count of the population reporting as any minority identity on the census
- `poverty` is an integer count of the population with household income below the poverty level

Let's map the `zones`.

```{r}
zones_map <- zones |> 
  tm_shape() + 
  tm_polygons(col = "grey30",
              lwd = 2,
              lty = "dashed",
              fill_alpha = 0) +
  tm_text("zone",
          col="grey30")  

pointgrid_map +
  zones_map
```

Let's load in the `parks`.

```{r}
parks <- st_read(here("data", "raw", "public", "cf_digitized.gpkg"), 
                     layer="parks")
```

Let's inspect the `parks` attribute table.

```{r}
parks |> st_drop_geometry() |> knitr::kable()
```

There is just one field, `park`, containing the park names.

Let's map the `parks`.

```{r}
working_color = "darkgreen"
parks_map <- parks |> 
  tm_shape() + 
  tm_polygons(col = working_color,
              lwd = 2,
              fill = working_color,
              fill_alpha = 0.5) +
  tm_text("park",
          col = working_color)  

pointgrid_map +
  parks_map
```

Let's load in the `watersheds`

```{r}
watersheds <- st_read(here("data", "raw", "public", "cf_digitized.gpkg"), 
                     layer="watersheds")
```

Let's inspect the `watersheds` attribute table.

```{r}
watersheds |> st_drop_geometry() |> knitr::kable()
```

Let's map the `watersheds`.

```{r}
working_color = "royalblue3"
watersheds_map <- watersheds |> 
  tm_shape() + 
  tm_polygons(col = working_color,
              lwd = 2,
              fill = working_color,
              fill_alpha = 0.2) +
  tm_text("name",
          col = working_color,
          ymod = -1,
          fontface = "italic")   

pointgrid_map +
  watersheds_map
```

