---
output: html_document
---

```{r include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Hello, Central Falls world

Let's read a reference point grid from the geopackage `cf_given` in the layer `pointgrid`.

As we read the data in, we'll recalculate x- and y- offsets for label placement in `tmap`.

```{r}
pointgrid <- st_read(here("data", "raw", "public", "cf_given.gpkg"), 
                     layer="pointgrid") |> 
  mutate(xOff = xOff / 4,
         yOff = yOff / -4,
         order = case_when(
           order == 0 ~ 0.5,
           .default = order
         ))
```

Let's display the grid.
We add a buffer to the pointgrid so that we have sufficient space surrounding the grid in which to place the labels.

```{r}
pointgrid_map <- pointgrid |> 
  tm_shape(bbox = st_buffer(pointgrid, 250)) +
  tm_symbols(border.alpha = 0.5,
             alpha = 0,
             size = "order",
             legend.size.show = FALSE) +
  tm_text("label",
          ymod="yOff",
          xmod="xOff",
          col="grey50")

pointgrid_map
```

Let's read a layer of Central Falls `zones`, the local geographic unit of enumeration and neighborhood governance.

```{r}
zones <- st_read(here("data", "raw", "public", "cf_digitized.gpkg"), 
                     layer="zones")
```

Let's inspect the `zones` attribute table.

```{r}
zones |> st_drop_geometry()
```

Notice how the `id` column has been converted into row labels.
There are four other columns:
- `zone` is character string with the zone name based on compass direction.
- `pop` is an integer count of the total population
- `minority` is an integer count of the population reporting as any minority identity on the census
- `poverty` is an integer count of the population with household income below the poverty level

Let's map the `zones`.

```{r}
zones_map <- zones |> 
  tm_shape() + 
  tm_polygons(border.col = "grey30",
              lwd = 2,
              lty = "dashed",
              alpha = 0) +
  tm_text("zone",
          col="grey30")  

pointgrid_map +
  zones_map
```

Let's load in the `parks`.

```{r}
parks <- st_read(here("data", "raw", "public", "cf_digitized.gpkg"), 
                     layer="parks")
```

Let's inspect the `parks` attribute table.

```{r}
parks |> st_drop_geometry()
```

There is just one field, `park`, containing the park names.

Let's map the `parks`.

```{r}
working_color = "darkgreen"
parks_map <- parks |> 
  tm_shape() + 
  tm_polygons(border.col = working_color,
              lwd = 2,
              col = working_color,
              alpha = 0.5) +
  tm_text("park",
          col=working_color,
          along.lines = TRUE)  

pointgrid_map +
  parks_map
```

Let's load in the `watersheds`

```{r}
watersheds <- st_read(here("data", "raw", "public", "cf_digitized.gpkg"), 
                     layer="watersheds")
```

Let's inspect the `watersheds` attribute table.

```{r}
watersheds |> st_drop_geometry()
```

Let's map the `watersheds`.

```{r}
working_color = "royalblue3"
watersheds_map <- watersheds |> 
  tm_shape() + 
  tm_polygons(border.col = working_color,
              lwd = 2,
              col = working_color,
              alpha = 0.2) +
  tm_text("name",
          col=working_color,
          along.lines = TRUE)   

pointgrid_map +
  watersheds_map
```