---
output: html_document
---

```{r include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Measuring shapes

Let's practice using spatial functions to measure and characterize the shapes of polygons using the example of `watersheds`.
All of these measurements are dependent upon the accuracy of the map projection in which the data is stored.

Also, different from normal GIS, measurements produce a value with a unit in R! 
This *can* be helpful, but when it isn't, it's easy enough to convert back to a simple number with `as.numeric()`.

## Area and Perimeter

What is the area of each watershed?
You'll need to know which column stores the geographic *geometry* data. 

```{r}
watersheds_shp <- mutate(watersheds, 
                     warea = st_area(geom),
                     wperim = st_perimeter(geom))

watersheds_shp |> st_drop_geometry() |> knitr::kable()
```

We can also calculate the compactness as a function of perimeter and area.

```{r}
watersheds_shp <- mutate(watersheds_shp, 
                     compact = round(
                       as.numeric(
                         (400 * pi * warea) / wperim^2),
                       2))

watersheds_shp |> st_drop_geometry() |> knitr::kable()
```

Do the compactness scores make sense?
Let's map them.

```{r}
pointgrid_map +
  watersheds_shp |> 
    tm_shape() +
    tm_polygons(col = "compact",
                alpha = 0.5,
                legend.show = FALSE) +
    tm_text("compact")
```

Looks good to me!

## What's the point?

We also have a few methods for converting polygons to points.

*Centroids* are located at the center of gravity of a polygon.

```{r warning=FALSE}
watersheds_centroids <- watersheds_shp |> st_centroid()
```
Here are the centroids:

```{r}
pointgrid_map +
  watersheds_map +
  tm_shape(watersheds_centroids) +
  tm_symbols(
    size = 1,
    col = "blue"
  )
```

Centroids sometimes fall outside of the polygon, which might produce undesirable results for cartography or topological analysis. In those cases `point_on_surface` may be better.

```{r warning=FALSE}
watersheds_pt_on_surf <- watersheds_shp |> st_point_on_surface()
```

Let's see the points on surface in relation to the centroids.

```{r}
pointgrid_map +
  watersheds_map +
  tm_shape(watersheds_centroids) +
  tm_symbols(
    size = 1,
    col = "blue"
  ) +
  tm_shape(watersheds_pt_on_surf) +
  tm_symbols(
    size = 0.5,
    col = "white"
  )
```

## Minimum bounding geometries

Some measures of gerrymandering rely on the ratio of the area to the area of the convex hull or of the minimum bounding circle.

Let's calculate the convex hull.

```{r warning = FALSE}
watersheds_hull <- watersheds_shp |> st_convex_hull()
```

And visualize the resulting polygons.

```{r}
pointgrid_map +
  watersheds_map +
  tm_shape(watersheds_hull) +
  tm_polygons(
    border.col = "white",
    col = "red",
    lwd = 3,
    lty = "dotted",
    alpha = 0.2
  ) 
```

How would the ratios of area to area of the convex hull come out?
The watersheds contributing to more overlapping areas whould result in lower compactness scores.

```{r}
watersheds_hull <- mutate(watersheds_hull,
                          compact_hull = round(as.numeric(warea / st_area(geom)), 2))

watersheds_hull |> knitr::kable()
```

Finally, let's find the minimum bounding circles and calculate the same.
https://r-spatial.github.io/lwgeom/reference/bounding_circle.html

```{r}
watersheds_boundcircle <- watersheds_shp |> st_minimum_rotated_()
```



```{r}
pointgrid_map +
  watersheds_map +
  tm_shape(watersheds_boundcircle) +
  tm_polygons(
    border.col = "white",
    col = "red",
    lwd = 3,
    lty = "dotted",
    alpha = 0.2
  ) 
```