---
output: html_document
---

# Area-weighted re-aggregation

Suppose that we need to summarize population data from `zones` in the boundaries of the `watersheds`.
We only have one commonality between the two layers: geographic location. 
To make matters more complicated, the boundaries are not neatly nested within one another.
In order to move data from the `zones` into the `watersheds`, we must assume that the population (and variables thereof) are equally distributed across the space of the zones.

## Calculate source areas

```{r}
zones <- mutate(zones, s_area = st_area(geom))
```


## Disaggregate

Let's use the `st_intersection` operation to overlay `zones` and `watersheds`.

```{r warning=FALSE}
fragments <- st_intersection(zones, watersheds)
```

Resulting in the table...

```{r}
fragments |> st_drop_geometry() |> knitr::kable()
```

and the map...

```{r}
pointgrid_map +
  fragments |> rownames_to_column(var = "id") |> 
    tm_shape() +
    tm_polygons(fill_alpha = 0.5) +
    tm_text("id")
```

Now, calculate area-weighted estimates.
We use an `as.numeric` function to discard the `units` data of `m^2`.

```{r}
fragments <- mutate(fragments,
                    f_area = st_area(geom),
                    aw = as.numeric(f_area / s_area),
                    aw_pop = aw * pop,
                    aw_minor = aw * minority,
                    aw_pov = aw * poverty)

fragments |> st_drop_geometry() |> knitr::kable(digits = 2)
```

## Reaggregate

Now it's time to re-aggregate into watersheds.
In R, we accomplish this by grouping with `group_by` and summarizing with `summarize`.
We dissolve the polygons together with `st_union`.
**Based on this, it may note be required to explicitly specify `geom = st_union(geom)`**
https://bookdown.org/robinlovelace/geocompr/geometric-operations.html


```{r}
watersheds_pop <- fragments |> group_by(name) |> 
  summarize(
    sumpop = sum(aw_pop),
    summinor = sum(aw_minor),
    sumpov = sum(aw_pov),
    geom = st_union(geom)
  )
```

Let's calculate the percentage minority and percentage below poverty and view the resulting attribute table.

```{r}
watersheds_pop <- mutate(watersheds_pop,
                         pctminority = round(summinor / sumpop * 100, 1),
                         pctpov = round(sumpov / sumpop * 100, 1))

watersheds_pop |> st_drop_geometry() |> knitr::kable(digits = 2)
```

And map the percentage in poverty in watersheds.

```{r}
pointgrid_map +
  watersheds_pop |> 
    tm_shape() +
    tm_polygons(fill = "pctminority",
                fill_alpha = 0.5,
                fill.legend = tm_legend_hide()) +
    tm_text("pctminority")
```
Another approach could have summarized the attribute data *without* the geometry data, and then joined the summaries back to the watersheds by attribute.

First, drop the geometry data and summarize the attribute data.

```{r}
watersheds_sum <- fragments |> 
  st_drop_geometry() |> 
  group_by(name) |> 
  summarize(
    sumpop = sum(aw_pop),
    summinor = sum(aw_minor),
    sumpov = sum(aw_pov)
  )

watersheds_sum |> knitr::kable(digits = 1)
```

Next, join the data to watersheds by attribute.

```{r}
watersheds_pop <- watersheds |> 
  left_join(watersheds_sum, by = join_by(name)) |> 
  mutate(pctminority = round(summinor / sumpop * 100, 1),
         pctpov = round(sumpov / sumpop * 100, 1))

watersheds_pop |> st_drop_geometry() |> knitr::kable(digits = 1)
```

And map the percentage of people in poverty in watersheds.

```{r}
pointgrid_map +
  watersheds_pop |> 
    tm_shape() +
    tm_polygons(fill = "pctminority",
                fill_alpha = 0.5,
                fill.legend = tm_legend_hide()) +
    tm_text("pctminority")
```
