---
output: html_document
---

# Projections and geodesic calculations

If only every study fit within a state plane zone or UTM zone... but alas, they do not.

How can we check the coordinate reference system and transform coordinates into another system?
How can we calculate area and length geodesically, rather than planimetrically?

## Projections

Let's check the projection of the watersheds layer.

```{r}
watersheds_crs <- st_crs(watersheds)

watersheds_crs$input
```

And let's see what the geometry data looks like.

```{r}
watersheds |> st_astext()
```
Now let's transform the data into NAD 1983 geographic coordinates.

```{r}
watersheds_NAD83 <- watersheds |> st_transform(4269)
```

And check the CRS.

```{r}
watersheds_NAD83 |> st_crs()
```

And inspect the geometry data to see if the coordinates have changed.

```{r}
watersheds_NAD83 |> st_astext()
```
Indeed, they are now stored as latitude and longitude coordinates!

## Geodesic calculations

As it turns out, *sf* will calculate spherical lengths and areas with `st_area` and `st_length` *if* the geometries are stored in a geographic coordinate system.
This is tested with the `st_is_longlat()` function.
Furthermore, there is a global setting for sf to switch spherical calculations off and use more accurate ellipsoidal calculations. This can be achieved with `sf_use_s2(FALSE)`. 

For perimeter calculations below, `st_perimeter(geom)` should work, and does work when running code blocks manually. 
For reasons I cannot explain, the operation needs to be simplified to a length function on a line in order for me to compile this code into a book.


```{r message=FALSE}
sf_use_s2(TRUE)

watersheds_NAD83 <- watersheds_NAD83 |>  mutate(
  spherical_area = st_area(geom),
  spherical_perim = st_length(st_cast(st_cast(geom, "POLYGON"), "LINESTRING"))
)

sf_use_s2(FALSE)

watersheds_NAD83 <- watersheds_NAD83 |>  mutate(
  ellipsoidal_area = st_area(geom),
  ellipsoidal_perim = st_length(st_cast(st_cast(geom, "POLYGON"), "LINESTRING"))
)

watersheds_NAD83 |> st_drop_geometry() |> kable()
```


Finally, you cannot calculate geodestic distance and area with geometries using projected coordinate systems in *sf*.
You could achieve this by transforming temporarily as you calculate, e.g. `st_area(st_transform(geom, 4269))`.
For example, we can calculate the planar and ellipsoidal areas of watersheds as follows:

```{r message=FALSE}
sf_use_s2(FALSE)

watersheds_measure <- watersheds |> 
  mutate(planar_area = st_area(geom),
         ellipsoidal_area = st_area(st_transform(geom, 4269)))

watersheds_measure |> st_drop_geometry() |> kable()
```